On the number of tiles touched by a\\ line segment on a rectangular grid

### 2021-08-02 paper.tex
Improved figures and some text

Given the coordinates $(x_1, y_1)$ of the first segment endpoint, with $0 \leq x_1 \leq a$, $0 \leq y_1 \leq b$, the second endpoint $(x_2, y_2)$ lies on a circle with radius $\len$ centered at $(x_1,y_1)$, as shown in Figure~\ref{fig: Pr_i}. The segment orientation is a random angle $\orient$ uniformly distributed on $[0,2\pi)$. It is clear from the figure that $i \geq n$ if and only if either $x_2 \geq a(n-1)$ or $x_2 \leq -a(n-2)$. The latter two events are exclusive, and by symmetry have the same probability. Therefore
\begin{equation}
\label{eq: Pr i = 2 Pr}
\Pr[i \geq n] = 2 \Pr[x_2 \geq a(n-1)].
\end{equation}

\begin{figure}
\centering%
\subfigure[$i$ can be equal or greater than $n$ for all $x_1$, $0 \leq x_1 \leq a$]{%
\label{fig: Pr_i_full}%
\includegraphics[width=.48\textwidth]{Pr_i_full_1p35}%
}\hfill%
\subfigure[$i$ can be equal or greater than $n$ only for $a(n-1)-\len \leq x_1 \leq a$]{%
\label{fig: Pr_i_notfull}%
\includegraphics[width=.48\textwidth]{Pr_i_notfull_1p35}%
}%
\caption{Conditions for the normalized width of the canonical rectangle, $i$, being equal or greater than a given $n$. Example with $a=1.35$, $b=1$
}%
\label{fig: Pr_i}
\end{figure}%

Consider the event $x_2 \geq a(n-1)$. There are three possibilities depending on $x_1$, $n$ and $\len$. If $a(n-1) < \len$, the length $\len$ is enough for $x_2$ to exceed $a(n-1)$ for some angles $\orient$, regardless of $x_1$. This is depicted in Figure~\ref{fig: Pr_i_full}, where the section of the arc with solid line represents those angles for which $x_2 > a(n-1)$. If $\len \leq a(n-1) < \len+a$, the length is enough only if $x_1 \geq a(n-1)-\len$, which corresponds to the shaded region in Figure~\ref{fig: Pr_i_notfull}. Lastly, if $\len+a \leq a(n-1)$ it is impossible that $x_2$ exceeds $a(n-1)$. Note that the value $y_1$ is irrelevant for this.

In the first two cases above, the probability that $x_2 \geq a(n-1)$, conditioned on $x_1$, is the length of the arc to the right of the line $x=a(n-1)$ divided by $2\pi\len$; that is,
\begin{equation}
\label{eq: Pr x_2 geq a(n-1) cond}
\Pr[x_2 \geq a(n-1) \cond x_1] = \frac 1 \pi \arccos \frac{a(n-1)-x_1}{\len}.
\end{equation}
In the first case $x_1$ has a uniform distribution on $(0,a)$, and $\Pr[x_2 \geq a(n-1)]$ is easily obtained from \eqref{eq: Pr x_2 geq a(n-1) cond} as follows:
\begin{equation}
\Pr[x_2 \geq a(n-1)] = \frac 1 {\pi a} \int_0^a\arccos \frac{a(n-1)-x_1}{\len} \diff x_1. 
\end{equation}
Integrating by parts, this is seen to be
\begin{equation}
\label{eq: Pr x_2 geq a(n-1), first case}
\Pr[x_2 \geq a(n-1)] = \frac{\len}{\pi a}\,\, \left( g\left( \frac{a(n-1)}{\len} \right) - g\left( \frac{a(n-2)}{\len} \right) \right),
\end{equation}
where the function $g$ is defined in \eqref{eq: g}. Substituting into \eqref{eq: Pr i = 2 Pr} yields the result in \eqref{eq: Pr i geq n}, second line.

@@

Secondly, a grid with $b \rightarrow \infty$ corresponds to Buffon's arrangement of parallel strips with spacing $a$. Thus
\begin{equation}
\label{eq: number of lines crossed, Buffon}
\lim_{b \rightarrow \infty} (\funta(\len)-1) = \frac{2\len}{pi a})
\end{equation}
gives the average number of lines crossed in Buffon's experiment. For $\len \leq a$ the segment can only cross one line, and \eqref{eq: number of lines crossed, Buffon} coincides with the probability of crossing \cite[section 1.1]{Mathai99}.

@@

***Maybe a graph comparing maximum and average. Or compute/plot their ratio. The asymptotic slopes are $\sqrt(1/a^2+1/b^2)$ for $\funt$ and $2/\pi(1/a+1/b)$. It may be interesting to consider the ratio of asymptotic slopes. Some quick plot of that ratio suggests it is maximum for $a=b$, and equals $sqrt(8)/pi = 0.9003$. The minimum number of tiles that can be touched with non-zero probability (corresponding to an n times 1 canonical rectangle if a>=b) can also be plotted.

>>>>

Given the coordinates $(x_1, y_1)$ of the first segment endpoint, with $0 \leq x_1 \leq a$, $0 \leq y_1 \leq b$, the second endpoint $(x_2, y_2)$ lies on a circle with radius $\len$ centered at $(x_1,y_1)$, as shown in Figure~\ref{fig: Pr_i}. The segment orientation is a random angle $\orient$ uniformly distributed on $[0,2\pi)$. It is clear from the figure that $i \geq n$ if and only if $x_2 \geq a(n-1)$ or $x_2 \leq -a(n-2)$; and these events are exclusive for $n \geq 2$. Thus
\begin{equation}
\label{eq: Pr i cond}
\Pr[i \geq n \cond x_1] = \Pr[x_2 \geq a(n-1) \cond x_1] + \Pr[x_2 \leq -a(n-2) \cond x_1].
\end{equation}
The two conditional probabilities on the right-hand side of \eqref{eq: Pr i cond} are different in general. However, averaging over $x_1$ gives, by symmetry, $\Pr[x_2 \geq a(n-1)] = \Pr[x_2 \leq -a(n-2)]$; and consequently, for $n \geq 2$,
\begin{equation}
\label{eq: Pr i = 2 Pr}
\Pr[i \geq n] = 2 \Pr[x_2 \geq a(n-1)].
\end{equation}

\begin{figure}
\centering%
\subfigure[$x_2$ can exceed $a(n-1)$ for all $x_1$, $0 \leq x_1 \leq a$]{%
\label{fig: Pr_i_full}%
\includegraphics[width=.58\textwidth]{Pr_i_full_1p35}%
}\\%\hfill%
\subfigure[$x_2$ can exceed $a(n-1)$ only if $a(n-1)-\len \leq x_1 \leq a$]{%
\label{fig: Pr_i_notfull}%
\includegraphics[width=.58\textwidth]{Pr_i_notfull_1p35}%
}%
\caption{Conditions for the normalized width of the canonical rectangle, $i$, being equal or greater than a given $n$. Example with $a=1.35$, $b=1$
}%
\label{fig: Pr_i}
\end{figure}%

Consider the event $x_2 \geq a(n-1)$ conditioned on $x_1$, with $n \geq 2$. There are three possibilities depending on $x_1$, $n$ and $\len$. If $a(n-1) < \len$, regardless of $x_1$ the length $\len$ is enough for $x_2$ to exceed $a(n-1)$ for some angles $\orient$. This is depicted in Figure~\ref{fig: Pr_i_full}, where the section of the arc with solid line represents those angles for which $x_2 > a(n-1)$. If $\len \leq a(n-1) < \len+a$, the length can only be enough provided that $x_1 \geq a(n-1)-\len$, which corresponds to the shaded region in Figure~\ref{fig: Pr_i_notfull}. Lastly, if $\len+a \leq a(n-1)$ it is not possible for $x_2$ to exceed $a(n-1)$. Note that the coordinate $y_1$ is irrelevant to this.

In the first two cases above, the probability that $x_2 \geq a(n-1)$, conditioned on $x_1$, is the length of the arc to the right of the line $x=a(n-1)$ divided by $2\pi\len$, that is,
\begin{equation}
\label{eq: Pr x_2 geq a(n-1) cond}
\Pr[x_2 \geq a(n-1) \cond x_1] = \frac 1 \pi \arccos \frac{a(n-1)-x_1}{\len}.
\end{equation}
In the first case $x_1$ has a uniform distribution on $(0,a)$, and $\Pr[x_2 \geq a(n-1)]$ is easily obtained from \eqref{eq: Pr x_2 geq a(n-1) cond} as follows:
\begin{equation}
\Pr[x_2 \geq a(n-1)] = \frac 1 {\pi a} \int_0^a\arccos \frac{a(n-1)-x_1}{\len} \diff x_1. 
\end{equation}
Integrating by parts,
\begin{equation}
\label{eq: Pr x_2 geq a(n-1), first case}
\Pr[x_2 \geq a(n-1)] = \frac{\len}{\pi a}\,\, \left( g\left( \frac{a(n-1)}{\len} \right) - g\left( \frac{a(n-2)}{\len} \right) \right),
\end{equation}
where the function $g$ is defined in \eqref{eq: g}. Substituting into \eqref{eq: Pr i = 2 Pr} yields the result in \eqref{eq: Pr i geq n}, second line.

@@

Secondly, a grid with $b \rightarrow \infty$ corresponds to Buffon's arrangement of parallel strips with spacing $a$. Thus
\begin{equation}
\label{eq: number of lines crossed, Buffon}
\lim_{b \rightarrow \infty} \funta(\len)-1 = \frac{2\len}{\pi a}
\end{equation}
gives the average number of lines crossed in Buffon's experiment. For $\len \leq a$ the segment can cross at most one line, and \eqref{eq: number of lines crossed, Buffon} coincides with the probability of crossing \cite[section 1.1]{Mathai99}.

@@

***Maybe a graph comparing maximum and average. Or compute/plot their ratio. The asymptotic slopes are $\sqrt(1/a^2+1/b^2)$ for $\funt$ and $2/\pi(1/a+1/b)$. It may be interesting to consider the ratio of asymptotic slopes, as a function of $a/b$. It is maximum for $a=b$, and equals $2\sqrt 2/\pi = 0.9003$. It tends to $2/\pi$ for $a/b \rightarrow \infty$ and for $a/b \rightarrow 0$. The minimum number of tiles that can be touched with non-zero probability (corresponding to an n times 1 canonical rectangle if a>=b) can also be plotted.
